---
title: "Anova"
author: "Hans"
date: "9/21/2019"
output: html_document
---

# Setup - always run

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse)
p_load(nycflights13)
p_load(ggplot2)
p_load(Deriv)
p_load(deSolve)
p_load(phaseR)
p_load(gridExtra)
p_load(FSA)
theme_set(theme_minimal())
```

### --------------------------------------------------------------
### One-way anova, SAS example, pp. 155-156
### --------------------------------------------------------------


```{r Read and format Data}

Input =("
Location   Aam
Tillamook  0.0571 
Tillamook  0.0813 
Tillamook  0.0831 
Tillamook  0.0976 
Tillamook  0.0817 
Tillamook  0.0859 
Tillamook  0.0735 
Tillamook  0.0659 
Tillamook  0.0923 
Tillamook  0.0836 
Newport    0.0873 
Newport    0.0662 
Newport    0.0672 
Newport    0.0819 
Newport    0.0749 
Newport    0.0649 
Newport    0.0835 
Newport    0.0725 
Petersburg 0.0974 
Petersburg 0.1352 
Petersburg 0.0817 
Petersburg 0.1016 
Petersburg 0.0968 
Petersburg 0.1064 
Petersburg 0.1050 
Magadan    0.1033 
Magadan    0.0915 
Magadan    0.0781 
Magadan    0.0685 
Magadan    0.0677 
Magadan    0.0697 
Magadan    0.0764 
Magadan    0.0689 
Tvarminne  0.0703 
Tvarminne  0.1026  
Tvarminne  0.0956 
Tvarminne  0.0973 
Tvarminne  0.1039 
Tvarminne  0.1045
")

Data = read.table(textConnection(Input), header = TRUE)
 
# Specify the order of factor levels for plots and Dunnett comparison
Data = mutate(Data, Location = factor(Location, levels = unique(Location)))
 
# Produce summary statistics
Summarize(Aam ~ Location, data=Data, digits=3)
```

```{r Model}

# Fit the linear model and conduct ANOVA
model = lm(Aam ~ Location, data = Data)

anova(model)       # Produces type I sum of squares
summary(model)     # Produces r-square, overall p-value, parameter estimates
```

```{r}

# Checking assumptions of the model
hist(residuals(model), col="darkgray")
plot(fitted(model), residuals(model))

```

```{r}
# Simple box plots of values across groups
 
boxplot(Aam ~ Location,
        data = Data,
        ylab="aam / height",
        xlab="Location")

### Summarize the data frame (Data) into a table

p_load(Rmisc)   

Data2 = summarySE(data=Data, 
          "Aam", 
          groupvars="Location", 
          conf.interval = 0.95)

Tabla = as.table(Data2$Aam)          
rownames(Tabla) = Data2$Location

barplot(Tabla, 
        ylab="aam / height", 
        xlab="Location")
 
```

```{r}
library(ggplot2)

offset.v = -3     # offsets for mean letters
offset.h = 0.5

ggplot(Data2, 
       aes(x = Location, y = Aam, 
           ymax=0.12, ymin=0.0))  +
       geom_bar(stat="identity", fill="gray50",
           colour = "black", width = 0.7)  +
       geom_errorbar(aes(ymax=Aam+se, ymin=Aam-se), 
                     width=0.0, size=0.5, color="black")  +
       geom_text(aes(label=c("bc","c","a","bc","ab"),
                 hjust=offset.h, vjust=offset.v)) +              
       labs(x = "Sample location", 
            y = "aam / height")  +
       ## ggtitle("Main title") + 
       theme_bw()  +
       theme(panel.grid.major.x = element_blank(),
             panel.grid.major.y = element_line(colour = "grey80"),
             plot.title = element_text(size = rel(1.5), 
             face = "bold", vjust = 1.5),
             axis.title = element_text(face = "bold"),
             axis.title.y = element_text(vjust= 1.8),
             axis.title.x = element_text(vjust= -0.5),
             panel.border = element_rect(colour="black")
             )
```


# Exmple II

```{r}
females = rnorm(4, mean=200, sd=20)
males = rnorm(4, mean=120, sd=20)
kidz = rnorm(4, mean=380, sd=20)
pitchstudy = data.frame(females, males, kidz)
```

```{r}
sum((females - mean(females))^2)
sum(females^2) - (sum(females)^2)/4

```{r}
bigvector = c(females, males, kidz)
CF = (sum(bigvector))^2/length(bigvector)
total.ss = sum(bigvector^2)-CF
between.ss = (sum(females)^2)/4 + (sum(males)^2)/4 + (sum(kidz)^2)/4 - CF
within.ss = total.ss - between.ss

df.total = length(bigvector)-1
df.between = ncol(pitchstudy)-1
df.within = df.total-df.between

between.ms = between.ss/df.between
within.ms = within.ss/df.within
F.value = between.ms/within.ms
    
1 - pf(F.value,2,9)
```


```{r}


# Now more automatic
groups = c(rep("female",4),rep("male",4),rep("kidz",4))
pitchstudy = data.frame(c(1:12), groups, bigvector)
colnames(pitchstudy) = c("subjects","groups","bigvector")
summary(aov(bigvector ~ groups + Error(subjects), data = pitchstudy))
summary(pitchstudy)
class(pitchstudy$subjects)
pitchstudy$subjects = as.factor(pitchstudy$subjects)
summary(aov(bigvector ~ groups + Error(subjects), data=pitchstudy))


```


