---
title: "Assignment 2"
author: "Hans"
date: "9/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



# Decay function

```{r}
decay <- function(t, k, P, A){
  A * exp(-k*t) * sin (2*pi*t/P)
}

decay(1, 0.1, 4, 1)

decay(P=4, t=1, A=1, k=0.1)
```


# Question 1 and 2

```{r}
decay(P=4, t=0.1, A=0.5, k=2.0)
decay(P=4, t=100000, A=0.5, k=2.0)

```

# Question 3

```{r}
curve(decay(x, 0.1, 4, 1), 0, 50, xlab = "t", ylab = "decay", col = "red")
```
# Question 5

You can use a different variable than x, but then you have to declare it

```{r}
curve(decay(t, 0.1, 4, 1), 0, 50, xlab="t", ylab="decay", col="red", xname = 't')

```

# Question 6

Smoother version of the curve with 1000 points

```{r}
curve(decay(x, 0.1, 4, 1), 0, 50, xlab="t", ylab="decay", col="red", n = 1000)
```


# Question 7

```{r}
curve(decay(P=4, t, A=2, k=0.1), 0, 50, xlab="t", ylab="decay", col="blue", lty = 2, xname = 't', n = 1000)
curve(decay(P=4, t, A=1, k=0.1), 0, 50, xlab="t", ylab="decay", col="red", xname = 't', n = 1000, add = TRUE)
```

# Question 8

Different method of plotting

```{r}
xmin = 0
xmax = 50
xstep = 0.01

#define vector of x values and corresponding function values for plotting
xvals = seq(xmin, xmax, xstep) 
yvals1 = decay(P=4, xvals, A=2, k=0.1)
yvals2 = decay(P=4, xvals, A=1, k=0.1)

#create an empty plot
plot(1, xlab = "t", ylab = "decay", type = "n", xlim = c(xmin, xmax), ylim = c(min(yvals1), max(yvals1)))
 
lines(xvals, yvals1, ylab="decay", col="blue", lty=2)
lines(xvals, yvals2, ylab="decay", col="red")
```


# Question 9

```{r}
#install.packages("Deriv") 
library(Deriv)

f1 <- function(x, a, b, c) {
  a*x^2 + b*exp(x) + c
}
f1
(derf = Deriv(f1))


f2 <- function(x, a) {
  a*sin(b*x)
}
f2
(derf = Deriv(f2))

```


# Question 10

Now plot the function and its derivative

```{r}
a_val = 2
b_val = 3
c_val = 5
h_val = 0
#h_val = 100

f1 <- function(x, a, b, c, h) {
  a*x^2 + b*exp(x) + c - h
}

xmin = -10
xmax = 4
xstep = 0.01

#define vector of x values and corresponding function values for plotting
xvals = seq(xmin, xmax, xstep) 
yvals = f1(xvals, a_val, b_val, c_val, h_val)
  
#create an empty plot
plot(1, xlab = "t", ylab = "f1", type = "n", xlim = c(xmin, xmax), ylim = c(min(yvals), max(yvals)))
lines(xvals, yvals, ylab="f1", col="blue")


derf1 = Deriv(f1, "x")
derf1

#define vector of x values and corresponding function values for plotting
xvals = seq(xmin, xmax, xstep) 
dyvals = derf1(xvals, a_val, b_val, c_val, h_val)
plot(1, xlab = "t", ylab = "df1", type = "n", xlim = c(xmin, xmax), ylim = c(min(dyvals), max(dyvals)))
lines(xvals, dyvals, ylab="df1", col="red")

```


# Question 11

```{r}
xcross = uniroot(derf1, a = a_val, b = b_val, c = c_val, h = h_val, lower = -1, upper = 1)
#xcross
#xcross[[1]]
#(f1(xcross$root))
#class(cross)
#class(cross[1])
#(d = as.double(cross[1]))
#(val_at_minimum = f1(-0.469))
#(val_at_minimum = f1(d))
#(val_at_minimum = f1(as.double(xcross[1])))
xcross$root
(val_at_minimum = f1(xcross$root, a = a_val, b = b_val, c = c_val, h = h_val))

```


# Question 12

```{r}
a = 2
b = 3
c = 5

f1 <- function(x) {
  a*x^2 + b*exp(x) + c - 100
}

xcross = uniroot(f1, lower = -2, upper = 4)
xcross$root
(f1(xcross$root))
xcross = uniroot(f1, lower = -8, upper = -5)
xcross$root
(f1(xcross$root))
```


# Question 13

Now plot the function and its derivative

```{r}


f1 <- function(x, a, b, c) {
  a*x^2 + b*exp(x) + c
}

xmin = -5
xmax = 5
xstep = 0.01

#define vector of x values and corresponding function values for plotting
xvals = seq(xmin, xmax, xstep) 
yvals = f1(xvals, a = 2, b = 3, c = 5)
  
#create an empty plot
plot(1, xlab = "x", ylab = "f1(x)", type = "n", xlim = c(xmin, xmax), ylim = c(min(yvals), max(yvals)))
lines(xvals, yvals, ylab="f1", col="blue")


derf1 = Deriv(f1, "x")
derf1

#define vector of x values and corresponding function values for plotting
xvals = seq(xmin, xmax, xstep) 
dyvals = derf1(xvals, a = 2, b = 3, c = 5)
plot(1, xlab = "t", ylab = "df1", type = "n", xlim = c(xmin, xmax), ylim = c(min(dyvals), max(dyvals)))
lines(xvals, dyvals, ylab="df1", col="red")

(uniroot(derf1, lower = -5, upper = 3, a = 2, b = 3, c = 5))

```
# Question 14 Hill

```{r}
hill <- function(L,h) {
  L/(L+h)
}

lmin = 0
lmax = 50
step = 0.1

h = 5

#define vector of x values and corresponding function values for plotting
lvals = seq(lmin, lmax, step) 
hvals1 = hill(lvals, 5)
hvals2 = hill(lvals, 10)
  
#create an empty plot
plot(1, xlab = "L", ylab = "H(L)", type = "n", xlim = c(lmin, lmax), ylim = c(min(hvals1), max(hvals1)))
lines(lvals, hvals1, ylab="f1", col="blue")
lines(lvals, hvals2, ylab="f1", col="red")
```

