---
title: "Assignment 3"
author: "Hans"
date: "9/4/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!require("pacman")) +
  install.packages("pacman") 
library(pacman)
```



```{r}
p_load(deSolve)
```

Use is made of a special package, that requires very specific input. Not important to understand exactly how that works.
Just the line that starts with dY is important and that is the line that you change.


```{r}
expgrowth = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dY = a * Y
    list(c(dY)) 
  })
}

pars = c(a = 0.2)
inistate = c(Y = 1)
time_series = seq(0, 10, by = 0.1)

out = ode(y = inistate, times = time_series, func = expgrowth, parms = pars)
plot(out, main = "", ylab = "Y")
out
```


# Excercise 1

```{r}

expgrowth = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dY = a * Y
    list(c(dY)) 
  })
}

pars = c(a = -0.1)
inistate1 = c(Y = 1)
inistate2 = c(Y = 5)
inistate3 = c(Y = 10)
time_series = seq(0, 10, by = 0.1)


out1 = ode(y = inistate1, times = time_series, func = expgrowth, parms = pars)
out2 = ode(y = inistate2, times = time_series, func = expgrowth, parms = pars)
out3 = ode(y = inistate3, times = time_series, func = expgrowth, parms = pars)

# First plot three different graphs.
# Because out1 already contains all the data, you do not need to specify y data
plot(out1, main = "Y = 1", ylab = "Y", col = "red")
plot(out2, main = "Y = 5", ylab = "Y", col = "blue")
plot(out3, main = "Y = 10", ylab = "Y", col = "black")

# Now plot them in one plot
# create an empty plot
plot(1, xlab = "time", ylab = "Y ", type = "n", xlim = c(0, 10), ylim = c(min(out3), max(out3)))
 
# Because out1 already contains all the data, you do not need to specify y data
lines(out1, ylab = "Y = 1", col = "blue", lty=2)
lines(out2, ylab = "Y = 5", col = "red")
lines(out3, ylab = "Y = 10", col = "black")
```


# Exercise 2

```{r}
(finalstate = out[nrow(out),2])
```

# Exercise 3

```{r}
(finalstate = out[nrow(out),-1])

# Why does the following not work
#(finalstate = out[-1,-1])
```


# Exercise 4

```{r}
expgrowth = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dY = a * Y
    list(c(dY)) 
  })
}

pars = c(a = 0.1)
inistate1 = c(Y = 1)
time_series = seq(0, 10, by = 0.1)
out1 = ode(y = inistate1, times = time_series, func = expgrowth, parms = pars)

pars = c(a = -0.01)
inistate1 = out1[nrow(out1),-1]  # The start pointis where the first series stopped
time_series = seq(10, 100, by = 0.1)
out2 = ode(y = inistate1, times = time_series, func = expgrowth, parms = pars)

# Determine the minima and maxima - the values are in the 2nd column
(ymin = min(out1[,2], out2[,2]))
(ymax = max(out1[,2], out2[,2]))

#create an empty plot
plot(1, xlab = "time", ylab = "Y ", type = "n", xlim = c(0, 100), ylim = c(ymin, ymax))

# Draw the lines
lines(out1, ylab = "Y = 1", col="blue", lty=2)
lines(out2, ylab = "Y = 1", col="blue", lty=2)
```


# Exercise 5

Found out by trying: unstable for a > 0 and stable for a < 0 

```{r}
expgrowth = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dY = a * Y
    list(c(dY)) 
  })
}
```

# An unstable plot

```{r}

pars <- c(a = 0.2)
phaseport = phasePortrait(expgrowth, parameters = pars, ylim = c(-2, 2), points = 10, xlab = "Y", ylab = "dY/dt (a = 1)", state.names = c("Y")) 
grid() 
stab.expgrowth = stability(expgrowth, ystar = 0, parameters = pars, system = "one.dim", state.names = c("Y"))

```

# A stable plot
```{r}
pars <- c(a = -0.2)
phaseport = phasePortrait(expgrowth, parameters = pars, ylim = c(-2, 2), points = 10, xlab = "Y", ylab = "dY/dt (a = 1)", state.names = c("Y")) 
grid() 
stab.expgrowth = stability(expgrowth, ystar = 0, parameters = pars, system = "one.dim", state.names = c("Y"))

```



```


# Question 6

The ODE is  dx/dt = b + k1*x - k2*x

or written as dx/dt = b + (k1 - k2)*x

Stable equilibrium for k1 - k2 < 0
UnStable equilibrium for k1 - k2 > 0


```{r}
gene_reg = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dx = b + k1 * x - k2 * x
    list(c(dx)) 
  })
}

pars1 = c(b = 2, k1 = 0.5, k2 = 0.45)
pars2 = c(b = 2, k1 = 0.5, k2 = 0.7)
inistate = c(x = 1)
time_series = seq(0, 20, by = 0.1)

out1 = ode(y = inistate, times = time_series, func = gene_reg, parms = pars1)
out2 = ode(y = inistate, times = time_series, func = gene_reg, parms = pars2)

# Note that this is yet another way to plot. The 'plot' function accepts more than one dataset
plot(out1, out2, main = "", ylab = "X")

stab.expgrowth = stability(gene_reg, ystar = 0, parameters = pars1, system = "one.dim", state.names = c("x"))

stab.expgrowth = stability(gene_reg, ystar = 0, parameters = pars2, system = "one.dim", state.names = c("x"))
```



# Question 7


# Question 8


# Question 9


# Question 10

Create a slightly more sophisticated function gene_reg2
Notice that at chosen parameters the equilibrium is at 10

```{r}
gene_reg2 = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dx = b + k1 * x - k2 * x^2
    list(c(dx)) 
  })
}

pars = c(b = 5, k1 = 0.5, k2 = 0.1)
inistate = c(x = 1)
time_series = seq(0, 10, by = 0.1)

library(deSolve)
out2 = ode(y = inistate, times = time_series, func = gene_reg2, parms = pars)

plot(out2, main = "", ylab = "x")
```


# Question 11

Now change the starting condutions.

Notice that at each starting condition the equilibrium is at 10, but you see a difference in the shape.

```{r}

gene_reg2 = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dx = b + k1 * x - k2 * x^2
    list(c(dx)) 
  })
}

pars = c(b = 5, k1 = 0.5, k2 = 0.1)
time_series = seq(0, 10, by = 0.1)

for (state in c(0.1, 1, 10, 20, 30) ) {
  inistate = c(x = state)
  out2 = ode(y = inistate, times = time_series, func = gene_reg2, parms = pars)
  plot(out2, main = "", ylab = "x")
}




```

# Question 12

Now look at the Phase portrait.

The function is an inverted parabola, which crosses the x-as at -5 and 10. 
From the function you can see indeed that for every value above 0 tyhe equilibrium is at 10.


```{r}

library("phaseR")
gene_reg2 = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dx = b + k1 * x - k2 * x^2
    list(c(dx)) 
  })
}
pars = c(b = 5, k1 = 0.5, k2 = 0.1)

phaseport = phasePortrait(gene_reg2, parameters = pars, ylim = c(-10, 15), 
                          points = 10, xlab = "x", ylab = "dx/dt", state.names = c("x"))
```

You can also see that for in fact for values > -5 the equlibrium is at 10.
We run the code from question 11 again, but now with negative values also

```{r}
library(deSolve)

gene_reg2 = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dx = b + k1 * x - k2 * x^2
    list(c(dx)) 
  })
}

pars = c(b = 5, k1 = 0.5, k2 = 0.1)
time_series = seq(0, 10, by = 0.1)

for (state in c(1, 20) ) {
  inistate = c(x = state)
  out2 = ode(y = inistate, times = time_series, func = gene_reg2, parms = pars)
  plot(out2, main = "", ylab = "x")
}
```

If you enter a value less than -5 there is no solution

```{r}
library(deSolve)

gene_reg2 = function(timepoint, state, parameters) { 
  with (as.list(c(state, parameters)), {
    dx = b + k1 * x - k2 * x^2
    list(c(dx)) 
  })
}

pars = c(b = 5, k1 = 0.5, k2 = 0.1)
time_series = seq(0, 10, by = 0.1)

for (state in c(-5) ) {
  inistate = c(x = state)
  out2 = ode(y = inistate, times = time_series, func = gene_reg2, parms = pars)
  plot(out2, main = "", ylab = "x")
}
```

