---
title: "Data Wrangling"
author: "Hans"
date: "9/9/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse)
p_load(nycflights13)
```


# Question 1

There are two differences:

* The sequence of the columns
* The capitalisatiion of weight

Unclear why this works!

```{r}
# The data is in long format
cw_l1 <- ChickWeight %>% as_tibble

cw_l1

#
# This syntax is not understood
#

cw_w <- cw_l1 %>% spread(-Chick, -Diet)
cw_w

#
# This syntax is more clear
#

cw_w <-  cw_l1 %>% spread(key = 'Time', value = weight)
cw_w

#
# This syntax is not understood
#

cw_l2 <- cw_w %>% gather(key = "Time", value = "Weight", 
                         -Chick, -Diet) %>% arrange(Chick)
cw_l2

#
# This syntax is more clear
#

cw_l2 <- cw_w %>% gather(key = "Time", value = "Weight", c(`0`:`21`))
cw_l2


# You can arrange the output

cw_l2 %>%
  arrange(as.numeric(Time), desc(Diet))

# But for some reason you can not arrange on Chick
cw_l2 %>%
  mutate(seq_chick = as.numeric(Chick)) %>%
  arrange(seq_chick)

```

# Exercises with Iris


## Exercise 2

Iris is tidy. It is in long format

## Exercise 3

Iris can be made 'longer'

```{r}
head(iris, 20) 
iris <- iris %>% as_tibble

iris
iris_longer1 <- gather(iris, key = 'Variable', value = "Measurement",
                       Sepal.Length:Petal.Width)
iris_longer1

iris_longer2 <- gather(iris, key ='Variable', value = "Measurement",
                       c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width))
iris_longer2

iris_longer3 <- iris %>% gather(key ='Variable', value = "Measurement",
                                Sepal.Length:Petal.Width)
iris_longer3
```


```{r}
library(dplyr)
# From http://stackoverflow.com/questions/1181060
stocks <- tibble(
  time = as.Date('2009-01-01') + 0:9,
  X = rnorm(10, 0, 1),
  Y = rnorm(10, 0, 2),
  Z = rnorm(10, 0, 4)
)
stocks
gather(stocks, "stock", "price", X:Z)
stocks %>% gather("stock", "price", -time)
```

```{r}
head(iris,20)
iris_longer <- iris %>% gather(key = 'Variable', value ="Measurement",
                               c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width))
iris_longer
```


# Exercise 4

Now try to go backwards

That does not work: you get an error

```{r}
# iris_longer %>% spread(key = 'Variable', value = 'Measurement')
```

# Exercise 5

To circumvent the problem you need to add a row number

```{r}
iris
iris_longer <- iris %>% 
               rowid_to_column %>% 
               gather(key = 'Variable', value ="Measurement", 
                      c(Sepal.Length, Sepal.Width, Petal.Length, Petal.Width))
iris_longer
iris_wide <- iris_longer %>% spread(key = 'Variable', value = 'Measurement')
iris_wide
```

# Exercise 6
```{r}
il = iris_longer %>% separate(col = Variable, sep="[.]", into = c("Var 1", "Var 2")) %>% select(-rowid)
il
```

# Exercise 7 Make the tables look alike 

```{r}

# The iris_wide table looks like this
iris_wide

# It should be the same as the iris table
iris

# This is iris_wide after conversion
iris_wide %>% select(Sepal.Length,Sepal.Width, Petal.Length, Petal.Width, Species)
```

# Exercise 8 Flowers of setosa that have Sepal.Length bigger than 5.0

```{r}
filter(iris, Sepal.Length > 5, Species == "setosa") 
```


# Exercise 9 Flowers of setosa and versicolor

```{r}
filter(iris, Sepal.Length > 5, Species == "setosa" | Species == "versicolor")
filter(iris, Sepal.Length > 5, Species %in% c("setosa","versicolor"))
```

# Excercise 10

```{r Excercise 10}
iris_wide %>% summarise(meanSepalLength = mean(Sepal.Length),
                        meanSepalWidth = mean(Sepal.Width),
                        meanPetalLength = mean(Petal.Width),
                        meanPetalWidth = mean(Petal.Width))

#
# Get the mean values for the four parameters for the three species 
#

iris_longer %>%  
  group_by(Species, Variable) %>%
  summarise((mean(Measurement)))
  

#
# Get the mean values for the four parameters (ignoring the different species) 
#

iris_longer %>%  
  group_by(Variable) %>%
  summarise((mean(Measurement)))


t = iris_longer %>% filter(Variable == "Sepal.Length") %>% select(Measurement)
t
```


# Excercise 11

```{r Excercise 11}
#filter(iris, Sepal.Length > 5, Species == "setosa" | Species == "versicolor")`

```


# Excercise 13

```{r Excercise 13}

library(nycflights13)
flights

# Here you get all the flights executed on each day by Envoy Air 

flights %>% left_join(airlines) %>% 
  select(year, month, day, flight, tailnum, name) %>% 
  filter(name == 'Envoy Air')

# Here you get all the flights executed by Envoy Air on 1 January

flights %>% left_join(airlines, by = 'carrier') %>% 
  select(year, month, day, flight, tailnum, name) %>% 
  filter(name == 'Envoy Air', year == 2013, month == 1, day == 1)

# If you want to know how many flights Envoy Air has each day

flights %>% left_join(airlines) %>% 
  filter(name == 'Envoy Air') %>% 
  count(year, month, day, flight, tailnum, name) 

# If you want to know how many flights Envoy Air has each month in 2013

flights %>% left_join(airlines) %>% 
  filter(name == 'Envoy Air') %>% 
  count(year, month) 

# If you want to know how many flights Envoy Air has every day in June month in 2013

flights %>% left_join(airlines) %>% 
  filter(name == 'Envoy Air', month == 6) %>% 
  count(day) 

# If you want to know how many flights Envoy Air has in June month in 2013

flights %>% left_join(airlines) %>% 
  select(year, month, day, flight, tailnum, name) %>% 
  filter(name == 'Envoy Air', year == 2013, month == 6) %>% count()

flights %>% left_join(airlines) %>% 
  select(tailnum, flight, name) %>% 
  filter(name == 'Envoy Air')

airlines %>% right_join(flights) %>% 
  select(year, month, day, tailnum, name) %>% 
  filter(name == 'Envoy Air')


# Analyse the flights table

# Some planes have been used a lot in 2013

flights %>% 
  count(tailnum)

# The number of different planes used is

dif_planes <-flights %>% 
  count(tailnum) %>% 
  count() 

# The number of different planes for which date is available is much smaller

plane_info <- planes %>% count()

cat("There are", dif_planes$n, " planes refered to in the flights database, but there are only",  plane_info$n, "plane descriptions." )

# How many records are there in the flights table?
flights %>%
  count()

# What is the average departure delay on the origin airports
flights %>%
  group_by(origin) %>%
  summarise(delay=mean(dep_delay))

# What is the average arrival delay on the destination airports
flights %>%
  group_by(dest) %>%
  summarise(delay=mean(arr_delay,na.rm = TRUE))

# What is the average departure delay per month?
d <- flights %>%
  group_by(month) %>%
  summarise(delay=mean(dep_delay, na.rm=TRUE))
d

ggplot(data = d) + 
    geom_bar(mapping = aes(x=month))


# How manuy destinations are there?
flights %>%
  group_by(dest) %>%
  summarise() %>% count()

 
```

# Exercise 14

```{r}

# Flights with a tailnum fro which no info is available from 'planes'
planes

anti_join(flights, planes, by = "tailnum")

anti_join(planes, flights, by = "tailnum")
```

