---
title: "Assignment 6 - Statistics"
author: "Hans"
date: "9/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(dplyr)
```


# Exercise 1, 2, 3

## Exercise 1
* predictor variable: independet
* response variable: dependent
* outcome variable: dependent
* explanatory variable: independent
* x-variable: independent
* y-variable: dependent

```{r}
glimpse(warpbreaks)
```

## Exercise 2
* Cannot really tell, but likely
* warpbreaks independent: wool and tension
* warpbreaks dependent: breaks

## Exercise 3

* warpbreaks independent: breaks and tension
* warpbreaks dependent: wool

# Exercise 4 - Student



```{r}
plot(x = rt(xseq, df = 5), y = xseq)
```

# Exercise 5 

It is a cumulative distribution function of a student distribution
```{r}
xseq <- seq(0,1, length.out = 1000)
t_dens <- rt(1000, df = 5)

# The quartile function
plot(y = qt(xseq, df = 5), x = xseq)

# The quartile function with the axes reversed: the CDF function
plot(x = qt(xseq, df = 5), y = xseq)

plot(y = pt(xseq, df = 5), x = xseq)
plot(y = dt(xseq, df = 5), x = xseq)
```

```{r}
# For the normal distributiion
# The quartile function

xseq <- seq(0, 1, length.out = 1000)
plot(y = qnorm(xseq), x = xseq)


# The quartile function with the axes reversed: the CDF function
plot(x = qnorm(xseq), y = xseq)

x = seq (-5, 5, 0.1)
plot(y = pnorm(x), x = x)

plot(y = dnorm(xseq), x = xseq)

x = seq (-5, 5, 0.1)
y = dnorm(x)
p <- ggplot(data = data.frame(x, y))
p + geom_line(mapping = aes(x = x, y = y), 
                   size = 1)

```

# Exercixe 7


```{r}
x <-rnorm(100000, 12, 2) 
set.seed(123) 
result_list = alist()
```

# Exercise 10 

```{r}
hist(sleep$extra)
shapiro.test(sleep$extra)
```
The p-value is high, so likely normal


# Exercise 11 Fold Change Data

* Is not normal distributed, because tyhe regions from 0 to 1 and 1 to infinity are not symmetric
* Make the x-axis logarithmic 

```{r}
my_data <- data.frame(gene = 0.1 + runif(10000), 
                      control = 0.1 +runif(10000))
my_data$fold_change <- my_data$gene/my_data$control
my_data$fold_change

hist(my_data$fold_change)
abline(v=1, col = 'red', lwd=2)
shapiro.test(my_data$fold_change[0:4999])

logfold = log2(my_data$fold_change)
hist(logfold)
abline(v=0, col = 'red', lwd=2)
shapiro.test(logfold[0:4999])

```