---
title: "R Notebook"
output:
  html_document:
    number_sections: yes
    toc: yes
    toc_depth: 2
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse)
p_load(nycflights13)
p_load(ggplot2)
p_load(modelr)
p_load(Deriv)
p_load(deSolve)
p_load(phaseR)
p_load(gridExtra)
theme_set(theme_minimal())
```


# Constructing normal distributions 

```{r}

# Generate two normal sampes, with mu = 0, sd = 1 
rnorm(2)

# Generate two normal sampes, specified mu and sd 
rnorm(2, mean = 5, sd = 1)
rnorm(2, mean = 5, sd = 200)

# Generate 100000 samples
# Specifying the seed ensures repeatability  
# If you set the seed you get a reoproducile effect. DOn't di it for now
# set.seed(123)
x1 <-rnorm(10, mean = 12, sd = 2)
x2 <-rnorm(100, mean = 12, sd = 2)
x3 <-rnorm(1000, mean = 12, sd = 2)
x4 <-rnorm(10000, mean = 12, sd = 2)
x5 <-rnorm(100000, mean = 12, sd = 2)
x6 <-rnorm(1000000, mean = 12, sd = 2)


# Plot density distribution of the 100000 points
# Creating a 'theoretical' normal curve 
# With more points it becines perfect

par(mfrow=c(3,3))
plot(density(x1))
plot(density(x2))
plot(density(x3))
plot(density(x4))
plot(density(x5))
plot(density(x6))

par(mfrow=c(1,1))

```




# Example of sampling size

```{r Example ofsampling size }

x <-rnorm(1000000, mean = 12, sd = 2)
plot(density(x))

n = 5
x_s <- sample(x, n) 


# Plot the samples.
# The x values are the sample, the y values are set to 0.01
# Points are drawn as blue diamonds


points(x = x_s, 
       y = rep(0.01, n), 
       col = "blue", 
       pch = 5, 
       cex = 2)

# Draw blue dotted vertical lines at the sample points 
abline(v = x_s, col = 'blue', lwd = 2 , lty = 2)

# Draw a red solid vertical line at the mean of the sample
abline(v = mean(x_s), col = 'red', lwd = 4 , lty = 1)

# Draw two red vertical dotted lines at mu - 2*sd and mu + 2*sd
abline(v = c(mean(x_s)-2*sd(x_s), mean(x_s)+2*sd(x_s)), 
       col = 'red', lwd = 1 , lty = 2)
```
# Cumulative sampling


Note: ecdf (Empirical Cumulative Distribution Function)

Compute an empirical cumulative distribution function, with several methods for plotting, printing and computing with such an “ecdf” object.

```{r Cumulative sampling}

# Now draw 100 samples
# Calculate the difference between these samples of 3
# Calculate the difference between the sample mean and the population true mean

x <-rnorm(1000000, mean = 12, sd = 2)
population_mean = mean(x)

sample_size = 3
result_list = alist()
for (i in 1:100) {
  x_s <- sample(x, sample_size) # sample from our x
    sample_mean = mean(x_s)
    result_list[[i]] <- abs(sample_mean - population_mean)  
}

# The density distribution of these mean differences is approximateky normal
# The cex parameter sets the size of the font
title = paste("Distribution of sample of -", sample_size, " - mean deviations from true mean")
plot(density(unlist(result_list)), 
     main = title,
     cex.main = 0.7)

s = seq(-1, 5, 0.25)
plot(hist(unlist(result_list),breaks=c(s)), 
     main = title,
     cex.main = 0.7)

# The emperical cumulative distrubution curve takes the density data and plots it cumulative
# Draw a vertical line at 1 ad a horizontal line at 0.65
plot(ecdf(unlist(result_list)), 
     main = "Cumulative plot of mean",
     cex.main = 0.7, 
     cex = 0.5)
abline(h = 0.65, v = 1, col = 'blue')

```

## Sample Error of Mean

```{r Sample Error of Mean}
sample_size <- 3

# If you take the number of experiments high, you get a nice approximation of the normal
# distribution (i.e. the histogram will resemble the curve)

repeat_experiment <- 10000     

# Generate 'sample_size * repeat_experiment' and put them in a matrix

population_sd = 2
population_mean = 12

sample_data <- matrix(rnorm(repeat_experiment * sample_size,  # number of sample
                            population_mean,                  # mean
                            population_sd),                   # sd   
                      repeat_experiment)
head(sample_data)
dim(sample_data)

# Of each sample calculate the mean.
# Plot these points in a histogram
sample_data_means <- rowMeans(sample_data)
head(sample_data_means)
hist(sample_data_means, freq = F)

# Overlay it with a 'theoretical' normal distriubution, use a high repeat to get a 
# perfect curve
abline(v = population_mean, col = 'red', lwd=4)
x <-rnorm(1000000, mean = 12, sd = population_sd / sqrt(sample_size))
lines(density(x))

sd(sample_data_means)
sem = 2 / sqrt(sample_size)
sem
```


# Manual ECDF from quantiles


```{r Manual ECDF}

n = 1000

# Creat a dataframe with 1 column with 1000 normal samples
# The samples are sorted

ndist = sort(rnorm(n))
showQuantiles <- data.frame(rnorm = ndist)

# Add a column to the dataframe
# The column contains a seq of 1000 points divided by 1 and rounded to 2 decimals 
# Simply by refering to a non-existent colums name you create it

showQuantiles$quantile <- round(seq(0,
                                    1, 
                                    length.out = nrow(showQuantiles)), 
                                digits = 2)

# You get a bunch of points with 0, a bunch of points at 0.01, at 0.02 etc 

# The curve is plotted
plot(x = showQuantiles$quantile, 
     y = showQuantiles$rnorm, 
     main = paste0("n = ", n ))


# The Cumulative Distribution Function is constructed by swapping the axes 
plot(y = showQuantiles$quantile, 
     x = showQuantiles$rnorm, 
     main = paste(paste0("n = ", n ), "ECDF"))

# Would have been easier to use the ecdf function
plot(ecdf(ndist))


```




# Confidence intervals

```{r Confidence intervals}

# Formula for Confidence Interval = mu +- q(a) * SEM
# Probably you have seen 'mean +- 1.96 + SEM'? That is the Z quantile for CL = 5%

qnorm(0.975)

# Generate 100 samples with specified mu and sd
# Normally speaking you do not know the population distribution. This is just for reference. 
n_sample = 100
pop_mean = 12
pop_sd = 2

# Take the samples and calculate the mean, sd and sem
x_s <- rnorm(n_sample, pop_mean, pop_sd)
sample_mean <- mean(x_s)
sample_sd <- sd(x_s)
sem = sample_sd / sqrt(n_sample)

# For fun, calculate mean and SD yourself to check
sn = length(x_s)
sm = sum(x_s) / sn
sd = sqrt(sum((x_s - sm)^2) / (sn - 1))


# Formula for confidence interval of 0.95
# The formula to calulate the quartile you need is 0.95 + (1 - 0.95) / 2
# With that new number you get the single sided quantile that leaves a right tail that is half 
# of what you need. 
# But then you need it left and right.........
dq = 0.95
q_to_use = dq + (1 - dq) / 2 

if (n_sample < 30) {  # Use the 'T' distrubution
  ci <- qt(q_to_use, df = n_sample - 1) * sem 
} else {              # Use the normal distrubution
  ci <- qnorm(q_to_use, sample_mean, sample_sd) * sem 
}

# Create a boxplot for the samples using base R
# boxplot(x_s, ylim = c(5,25))
# abline(h = c(sample_mean - ci,  sample_mean + ci), col = 'red', lty=2)
# abline(h = c(sample_mean - sample_sd, sample_mean + sample_sd), col = 'blue', lty=2)

# Create a boxplot for the samples using ggplot
ggplot(data = data.frame(sample = x_s)) +
  geom_boxplot(mapping = aes(y=sample)) +
  ylim(5, 25) +
  theme(axis.text.x = element_blank()) +
  geom_hline(yintercept = c(sample_mean - ci, sample_mean + ci), 
             col= 'red', 
             linetype = 'dashed') +
  geom_hline(yintercept = c(sample_mean - sample_sd, sample_mean + sample_sd), 
             col = 'blue', 
             linetype = 'dashed')

```




# Central limit theorem

```{r Central limit theorem}

# Generate a 100 sample normal sample distribution, three times 
n = 100
replicate(3, hist(runif(n), cex.main = 0.8))

# Repeat a 100 sample, 1000 times
# Every column is one sample
sample_set = replicate(1000, runif(n)) 

# Take the means of each of the 1000 samples
means = colMeans(sample_set)

# Plot that as a hist=ogram
hist(means, cex.main = 0.8)
```




## Bootstrapping

```{r}
# Create a sample of 5 numbers, values between 5 and 6
n = 5
x <- runif(n, min = 5, max = 6)  

# Plot the one sample
hist(x, main = "one sample", cex.main = 0.8)

# From the sample, select 5 times a number
# We replace, so you may find in a set of 5 the same numbers multiple times
# Do this experiment 1000 times
b <- replicate(10000, mean(sample(x, replace=TRUE)))


hist(replicate(10000, runif(n, min = 5, max =6)), 
     main = "uniform distribution", cex.main = 0.8)

hist(b, main = "bootstrapped" , cex.main = 0.8)

plot(ecdf(b), main = "bootstrapped - cdf" , cex.main = 0.8)

qb <- quantile(b, c(0.025,0.975))


```



# Testing for normality-transformations

The null-hypothesis of this test is that the population is normally distributed. Thus, on the one hand, if the p value is less than the chosen alpha level, then the null hypothesis is rejected and there is evidence that the data tested are not normally distributed. On the other hand, if the p value is greater than the chosen alpha level, then the null hypothesis that the data came from a normally distributed population can not be rejected (e.g., for an alpha level of .05, a data set with a p value of less than .05 rejects the null hypothesis that the data are from a normally distributed population)

```{r Shapiro Test}

# A uniform distribution is not expected to meet the shapiro test and generates 
# a very low p value
hist(runif(10000))
shapiro.test(runif(1000))
qqnorm(runif(1000))

# A uniform distribution is not expected to meet the shapiro test and generates 
# a high p value
hist(rnorm(10000))
shapiro.test(rnorm(1000))
qqnorm(rnorm(1000))

```

# Log FC transformation

```{r Log transformation}
# Generate a data set with a lot of values between 0 and 1, but also with values larger than 1
my_data <- data.frame(gene = 0.1 + runif(10000), 
                      control = 0.1 + runif(10000))
my_data$fold_change <- my_data$gene / my_data$control

# If you plot it in a normal histogram, it is not so easy to inspect because the high numbers
# determine the scale
hist(my_data$fold_change)
abline(v = 1, col = 'red', lwd=2)

# If you take a log of the data tham the high numbers become a lot smaller
hist(log2(my_data$fold_change))
abline(v = 0, col = 'red', lwd=2)
```

# Standard Normal
```{r Standard normal}

sample = rnorm(1000, 12, 2)
sample_mean = mean(sample)
sample_sd = sd(sample)
par(mfrow=c(1,2))
hist(sample, cex.main = 0.8)
hist((sample - sample_mean) / sample_sd, cex.main = 0.8)
par(mfrow=c(1,1))


```

