---
title: "Assignment 6 - Statistics"
author: "Hans"
date: "9/11/2019"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(pacman)
p_load(tidyverse)
p_load(ggplot2)
p_load(gridExtra)
theme_set(theme_minimal())
```


# Exercise 1, 2, 3

## Exercise 1
* predictor variable: independent
* response variable: dependent
* outcome variable: dependent
* explanatory variable: independent
* x-variable: independent
* y-variable: dependent

```{r}
glimpse(warpbreaks)
```

## Exercise 2
* Cannot really tell, but likely
* warpbreaks independent: wool and tension
* warpbreaks dependent: breaks

## Exercise 3

* warpbreaks independent: breaks and tension
* warpbreaks dependent: wool

# Exercise 4 - Student



# ```{r}
# plot(x = rt(xseq, df = 5), y = xseq)
# ```

# Exercise 5 

It is a cumulative distribution function of a student distribution
```{r}
xseq <- seq(0,1, length.out = 1000)
t_dens <- rt(1000, df = 5)

# The quartile function
plot(y = qt(xseq, df = 5), x = xseq)

# The quartile function with the axes reversed: the CDF function
plot(x = qt(xseq, df = 5), y = xseq)

plot(y = pt(xseq, df = 5), x = xseq)
plot(y = dt(xseq, df = 5), x = xseq)
```

```{r}
# For the normal distributiion
# The quartile function

xseq <- seq(0, 1, length.out = 1000)
plot(y = qnorm(xseq), x = xseq)


# The quartile function with the axes reversed: the CDF function
plot(x = qnorm(xseq), y = xseq)

x = seq (-5, 5, 0.1)
plot(y = pnorm(x), x = x)

plot(y = dnorm(xseq), x = xseq)

x = seq (-5, 5, 0.1)
y = dnorm(x)
p <- ggplot(data = data.frame(x, y))
p + geom_line(mapping = aes(x = x, y = y), 
                   size = 1)

```

# Exercise 6

```{r}
x <-rnorm(100000, 12, 2)
density(x)

nsamples = 1000
set.seed(123) # make random part reproducable

# Using more efficient code than in lectures, and use ggplot ;) sample_sizes_1 = 3

sample_sizes_1 = 3
sample_sizes_2 = 15
sim_data_1 <- matrix(rnorm(nsamples * sample_sizes_1, 12, 2), nsamples) 
sim_data_2 <- matrix(rnorm(nsamples * sample_sizes_2, 12, 2), nsamples) 
sim_data_means_1 <- rowMeans(sim_data_1)
sim_data_means_2 <- rowMeans(sim_data_2)

# Use neat tibble trick (do you understand what it does?):
combined_sim_data <- tribble(
  ~sample_size, ~data,
  sample_sizes_1, sim_data_means_1,
  sample_sizes_2, sim_data_means_2,
) %>% unnest()

combined_sim_data %>%
  ggplot(aes(data)) + 
  geom_histogram() + 
  geom_vline(xintercept = 12, color = "red") + 
  facet_wrap(vars(sample_size), labeller = label_both) + 
  labs(x = "Sample mean", title = "Distribution of sample means")

combined_sim_data %>%
  ggplot(aes(abs(data - 12))) + 
  geom_density() + 
  facet_wrap(vars(sample_size), labeller = label_both) + 
  labs(x = "Absolute difference in sample and real mean", 
       title = "Distribution of errors of sample means")


combined_sim_data %>%
  ggplot(aes(abs(data - 12))) + 
  stat_ecdf() + 
  geom_vline(xintercept = 1, color = "red") + 
  facet_wrap(vars(sample_size), labeller = label_both) + 
  labs(x = "Absolute difference in sample and real mean", 
  title = "Distribution of errors of sample means")

```


# Exercixe 7


```{r}
set.seed(123) 
x <-rnorm(500, mean = 5, sd = 2) 
par(mfrow=c(1,2))
plot(density(x))
abline(v=5)
plot(ecdf(x))
abline(v=5)
abline(h=0.5)
par(mfrow=c(1,1))
```

# Exercise 8

```{r}
warpbreaks
woolA = filter(warpbreaks, wool == "A")

h_mean  = mean(woolA[,"breaks"])
h_sd    = sd(woolA[,"breaks"])
h_n     = length(woolA[,"breaks"])
h_t_val = qt(0.975, df = h_n-1 )
h_sem   = h_sd/sqrt(h_n)
h_ci_p  = h_mean + h_t_val * h_sem
h_ci_m  = h_mean - h_t_val * h_sem

woolA$Index = seq.int(nrow(woolA))
ggplot(data = woolA) +
  geom_point(mapping = aes(x = Index, y = breaks)) +
  geom_hline(yintercept = h_mean, col = 'green') +
  geom_hline(yintercept = h_ci_p , col = 'red') +
  geom_hline(yintercept = h_ci_m, col = 'red') +
  scale_y_continuous(breaks=seq(10, 60, 10)) 
  

mean_woolA <- mean(warpbreaks$breaks[warpbreaks$wool  == "A"])
sd_woolA <- sd(warpbreaks$breaks[warpbreaks$wool  == "A"])
n <- sum(warpbreaks$wool  == "A")
t_val <- qt(0.95, df = n-1 )
error <- t_val * sd_woolA / sqrt(n)
par(mfrow=c(1,1))
plot(warpbreaks$breaks[warpbreaks$wool  == "A"])
abline(h = mean_woolA)
abline(h = c(mean_woolA - error, mean_woolA + error), lty = 2)
 
```


# Exercise 9

```{r}


# Split out commands to understand
s = sample(sleep$extra, 5, replace = TRUE)
m = mean(s)
ms = mean(sample(sleep$extra, 5, replace = TRUE))


# Take 5 samples, repeat this sampling 10000 times each time calculating the sample mean.
b <- replicate(10000, mean(sample(sleep$extra, 5, replace = TRUE))) 

# plot the ecdf of the means
plot(ecdf(b), main = "bootstrapped - cdf", cex.main = 0.8) 
plot(density(b), main = "bootstrapped - cdf", cex.main = 0.8) 
quantile(b, c(0.01, 0.99))
 
```

# Exercise 10 

```{r}
hist(sleep$extra)
shapiro.test(sleep$extra)
qqnorm(sleep$extra)
qqline(sleep$extra)
```
The p-value is high, so likely normal


# Exercise 11 Fold Change Data

* Is not normal distributed, because the regions from 0 to 1 and 1 to infinity are not symmetric
* Make the x-axis logarithmic 

```{r}
my_data <- data.frame(gene = 0.1 + runif(10000), 
                      control = 0.1 +runif(10000))
my_data$fold_change <- my_data$gene/my_data$control

hist(my_data$fold_change)
abline(v=1, col = 'red', lwd=2)
shapiro.test(my_data$fold_change[0:4999])

logfold = log2(my_data$fold_change)
hist(logfold)
abline(v=0, col = 'red', lwd=2)
shapiro.test(logfold[0:4999])

#The shapiro is still bad, butthat is maybe because of having to cutr off sample points to 5000
# Naybe with qqnorm ?

qqnorm(logfold) 

# Indeed, much better
```